---
title: "Getting Started with BiocRefgetStore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with BiocRefgetStore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

BiocRefgetStore provides a BSgenome-compatible interface to reference genomes
backed by GA4GH refget stores. Instead of managing FASTA files, you connect
to a refget store (local or remote) and access sequences by digest.

This tutorial uses the **2023 Human Pangenome Reference** -- a remote refget
store containing 47 haplotype-resolved assemblies hosted on S3. Sequences are
downloaded on-demand and cached locally, so you don't need to download the
entire dataset upfront.

## Installation

BiocRefgetStore depends on `gtars`, a Rust-backed R package. Because `gtars`
lives in a Rust monorepo with sibling crates, it must be installed from a local
clone of the full repository (not directly from GitHub):

```{r install, eval=FALSE}
# 1. Clone the gtars monorepo (if you haven't already)
# git clone https://github.com/databio/gtars.git

# 2. Install gtars from the local clone (requires Rust toolchain)
install.packages("path/to/gtars/gtars-r", repos = NULL, type = "source")

# 3. Install BiocRefgetStore
remotes::install_github("refgenie/refget", subdir = "refget-r", ref = "r")
# Or from local source:
# install.packages("path/to/refget/refget-r", repos = NULL, type = "source")
```

`gtars` is not on CRAN and `remotes::install_github()` won't work for it
because the R package depends on sibling Rust crates via relative paths.
You need the full monorepo checkout so those paths resolve correctly.

## Connect to a remote pangenome store

Load a genome from the Human Pangenome Reference store. The store metadata
(~1.5 MB) is fetched on first use; individual sequences are downloaded
on-demand and cached locally:

```{r remote-store, eval=FALSE}
library(BiocRefgetStore)

# 2023 Human Pangenome Reference (47 haplotype-resolved assemblies)
pangenome_url <- "https://refgenie.s3.us-east-1.amazonaws.com/pangenome_refget_store"

# One assembly from the pangenome (HG03540.pri.mat.f1_v2)
genome <- RefgetGenome.from_remote(
  cache_path = "~/.cache/refget/pangenome",
  remote_url = pangenome_url,
  digest = "0aHV7I-94paL9Z1H4LNlqsW3WxJhlou5"
)
genome
#> RefgetGenome with 750 sequences
#>   collection_digest: 0aHV7I-94paL9Z1H4LNlqsW3WxJhlou5
#>   seqnames: JAGYVX010000001.1, JAGYVX010000002.1, ... (745 more)
```

Subsequent calls reuse the local cache -- no re-downloading.

## Basic sequence access

Extract a full sequence or a region by coordinates:

```{r getseq, eval=FALSE}
# Full sequence (returns DNAString if Biostrings is installed)
seq <- genome[["JAGYVX010000001.1"]]

# Region by coordinates (1-based, inclusive)
region <- getSeq(genome, "JAGYVX010000001.1", start = 1000, end = 2000)

# Force character output
region_chr <- getSeq(genome, "JAGYVX010000001.1", start = 1000, end = 2000,
                     as.character = TRUE)
```

Negative-strand extraction applies the reverse complement:

```{r strand, eval=FALSE}
rc <- getSeq(genome, "JAGYVX010000001.1", start = 1000, end = 2000, strand = "-")
```

## Multiple regions at once

Pass vectors of names, starts, and ends:

```{r vectorized, eval=FALSE}
seqs <- getSeq(
  genome,
  names = c("JAGYVX010000001.1", "JAGYVX010000002.1", "JAGYVX010000003.1"),
  start = c(100, 200, 300),
  end   = c(199, 299, 399)
)
```

If you have a GRanges object, pass it directly:

```{r granges, eval=FALSE}
library(GenomicRanges)
gr <- GRanges(c("JAGYVX010000001.1:100-199:+", "JAGYVX010000002.1:200-299:-"))
seqs <- getSeq(genome, gr)
```

## Bulk extraction from a data.frame

`extractRegions` accepts a data.frame with `chrom`, `start`, `end` columns:

```{r extract-regions, eval=FALSE}
regions <- data.frame(
  chrom = c("JAGYVX010000001.1", "JAGYVX010000001.1", "JAGYVX010000002.1"),
  start = c(100, 5000, 200),
  end   = c(199, 5099, 299)
)
seqs <- extractRegions(genome, regions, as.character = TRUE)
```

## Export sequences to FASTA

Write extracted regions or full sequences to a FASTA file:

```{r export, eval=FALSE}
# Regions to FASTA
extractToFasta(genome, regions, "extracted_regions.fa")

# Specific sequences
exportChromosomes(genome, c("JAGYVX010000001.1", "JAGYVX010000002.1"), "subset.fa")

# All sequences
exportChromosomes(genome, output_path = "full_assembly.fa")
```

## Genome metadata

Inspect sequences and their properties:

```{r metadata, eval=FALSE}
seqnames(genome)       # sequence names
seqlengths(genome)     # named integer vector of lengths
seqinfo(genome)        # full Seqinfo object
length(genome)         # number of sequences
collection_digest(genome)   # seqcol digest
coordinate_system(genome)   # sorted_name_length_pairs digest
sequence_digests(genome)    # per-sequence SHA512t24u digests
```

## Working with local FASTA files

You can also create a genome directly from a local FASTA file. This builds
an in-memory refget store, computes sequence digests, and creates a `Seqinfo`
object automatically:

```{r from-fasta, eval=FALSE}
genome <- RefgetGenome.from_fasta("genome.fa")
genome
```

## Persistent on-disk store

For large genomes you access repeatedly, use an on-disk store so sequences
are indexed once and reused across sessions:

```{r on-disk, eval=FALSE}
# First time: create the store from FASTA
store <- gtars::refget_store_on_disk("~/.local/share/refget/hg38")
result <- gtars::add_fasta(store, "hg38.fa")
# Save the digest: result$digest

# Later: reload without re-parsing
genome <- RefgetGenome.from_directory(
  "~/.local/share/refget/hg38",
  digest = "saved_digest_string"
)
```

## Next steps

See the [Reference](reference.html) vignette for complete documentation of
every function and method in the package.
