---
title: "BiocRefgetStore Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BiocRefgetStore Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Complete reference for every exported function, method, and class in
BiocRefgetStore.

## Constructors

### RefgetGenome.from_fasta

```r
RefgetGenome.from_fasta(fasta_path)
```

Create a `RefgetGenome` from a FASTA file. Builds an in-memory refget store,
computes digests, and indexes all sequences.

- **fasta_path** — Path to a FASTA file (`.fa`, `.fasta`, `.fa.gz`).
- **Returns** — A `RefgetGenome` object.

```{r from-fasta, eval=FALSE}
genome <- RefgetGenome.from_fasta("hg38.fa")
```

### RefgetGenome.from_directory

```r
RefgetGenome.from_directory(path, digest = NULL, namespace = NULL, alias = NULL)
```

Load a `RefgetGenome` from a persisted on-disk refget store directory.

- **path** — Path to a directory created by `gtars::refget_store_on_disk()`.
- **digest** — Collection digest string (provide this OR namespace + alias).
- **namespace** — Alias namespace (e.g., `"refseq"`).
- **alias** — Alias name (e.g., `"GRCh38"`).
- **Returns** — A `RefgetGenome` object.

```{r from-directory, eval=FALSE}
genome <- RefgetGenome.from_directory("~/.refget/hg38", digest = "abc123...")
```

### RefgetGenome.from_remote

```r
RefgetGenome.from_remote(cache_path, remote_url, digest = NULL, namespace = NULL, alias = NULL)
```

Create a `RefgetGenome` backed by a remote refget store with local caching.

- **cache_path** — Local directory for caching downloaded data.
- **remote_url** — URL of the remote refget store.
- **digest** / **namespace** / **alias** — Same as `RefgetGenome.from_directory`.
- **Returns** — A `RefgetGenome` object.

```{r from-remote, eval=FALSE}
genome <- RefgetGenome.from_remote(
  cache_path = "~/.cache/refget",
  remote_url = "https://refget.databio.org/store",
  namespace = "refseq", alias = "GRCh38"
)
```

### RefgetGenome (low-level)

```r
RefgetGenome(store, digest = NULL, namespace = NULL, alias = NULL)
```

Construct a `RefgetGenome` from an existing `gtars::RefgetStore` object.
Requires either `digest` or both `namespace` and `alias`.

- **store** — A gtars `RefgetStore` object.
- **digest** / **namespace** / **alias** — Collection identifier.
- **Returns** — A `RefgetGenome` object.

```{r constructor, eval=FALSE}
store <- gtars::refget_store_open_local("/path/to/store")
genome <- RefgetGenome(store, namespace = "refseq", alias = "GRCh38")
```

---

## Sequence Access

### getSeq

```r
getSeq(x, names, start = NA, end = NA, strand = "+", as.character = FALSE, ...)
```

Extract sequences from a `RefgetGenome`. BSgenome-compatible interface.

- **x** — A `RefgetGenome` object.
- **names** — Character vector of sequence names, or a `GRanges` object.
- **start** — Integer start position(s), 1-based inclusive. `NA` for full sequence.
- **end** — Integer end position(s), 1-based inclusive. `NA` for full sequence.
- **strand** — `"+"` (default) or `"-"` for reverse complement.
- **as.character** — If `TRUE`, return character instead of DNAString/DNAStringSet.
- **Returns** — Single sequence: `DNAString` (or character). Multiple: `DNAStringSet` (or character vector). Named as `"seqname:start-end"` for regions.

```{r getseq, eval=FALSE}
# Full chromosome
getSeq(genome, "chr1")

# Region
getSeq(genome, "chr1", start = 100, end = 200)

# Reverse complement
getSeq(genome, "chr1", start = 100, end = 200, strand = "-")

# Multiple regions
getSeq(genome, c("chr1", "chr2"), c(100, 500), c(200, 600))

# From GRanges
getSeq(genome, GRanges("chr1:100-200:-"))
```

### `[[` (bracket extraction)

```r
genome[["chr1"]]
```

Extract a full sequence by name. Returns `DNAString` if Biostrings is
installed, otherwise a character string.

- **i** — Sequence name (character).
- **Returns** — `DNAString` or character string.
- **Errors** — If the sequence name is not found in the collection.

---

## Metadata Accessors

### seqinfo

```r
seqinfo(x)
```

Returns the `Seqinfo` object containing sequence names and lengths.

- **Returns** — A `GenomeInfoDb::Seqinfo` object.

### seqnames

```r
seqnames(x)
```

Returns the sequence names.

- **Returns** — Character vector (via `Seqinfo`).

### seqlengths

```r
seqlengths(x)
```

Returns named integer vector of sequence lengths.

- **Returns** — Named integer vector.

```{r seqlengths, eval=FALSE}
seqlengths(genome)
#>   chr1   chr2   chr3
#> 248956 242193 198295
```

### length

```r
length(x)
```

Returns the number of sequences in the genome.

- **Returns** — Integer scalar.

### names

```r
names(x)
```

Returns the sequence names as a character vector.

- **Returns** — Character vector.

### collection_digest

```r
collection_digest(genome)
```

Returns the GA4GH seqcol digest identifying this sequence collection.

- **genome** — A `RefgetGenome` object.
- **Returns** — Character string.

### coordinate_system

```r
coordinate_system(genome)
```

Returns the `sorted_name_length_pairs` digest. Two genomes with the same
`coordinate_system()` share the same coordinate system and are compatible for
coordinate-based operations (e.g., lifting over annotations).

- **genome** — A `RefgetGenome` object.
- **Returns** — Character string.

### sequence_digests

```r
sequence_digests(genome)
```

Returns a named character vector of per-sequence SHA512t24u digests.

- **genome** — A `RefgetGenome` object.
- **Returns** — Named character vector (names are sequence names, values are digests).

```{r seq-digests, eval=FALSE}
sequence_digests(genome)
#>                         chr1                          chr2
#> "SQ.2648ae1bacce4ec4b6cf337..." "SQ.f932a39b4c70..."
```

### store

```r
store(genome)
```

Returns the underlying `gtars::RefgetStore` object. Useful for calling gtars
functions directly.

- **genome** — A `RefgetGenome` object.
- **Returns** — A gtars `RefgetStore` object.

---

## Bulk Extraction

### extractRegions

```r
extractRegions(genome, regions, as.character = FALSE)
```

Extract multiple genomic regions efficiently using BED-based extraction.

- **genome** — A `RefgetGenome` object.
- **regions** — A `GRanges` object or a `data.frame` with columns `chrom`, `start`, `end` (1-based inclusive coordinates).
- **as.character** — If `TRUE`, return character vector instead of `DNAStringSet`.
- **Returns** — `DNAStringSet` or named character vector. Named as `"chrom:start-end"`.

```{r extract-regions, eval=FALSE}
regions <- data.frame(
  chrom = c("chr1", "chr1", "chr2"),
  start = c(100, 5000, 200),
  end   = c(199, 5099, 299)
)
seqs <- extractRegions(genome, regions)
```

### extractToFasta

```r
extractToFasta(genome, regions, output_path)
```

Write extracted regions directly to a FASTA file.

- **genome** — A `RefgetGenome` object.
- **regions** — A `GRanges` object or data.frame (same as `extractRegions`).
- **output_path** — Path for the output FASTA file.
- **Returns** — Invisibly returns `output_path`.

```{r extract-to-fasta, eval=FALSE}
extractToFasta(genome, regions, "output.fa")
```

### exportChromosomes

```r
exportChromosomes(genome, names = NULL, output_path, line_width = 80L)
```

Export complete chromosomes to a FASTA file.

- **genome** — A `RefgetGenome` object.
- **names** — Character vector of chromosome names to export, or `NULL` for all.
- **output_path** — Path for the output FASTA file.
- **line_width** — Bases per line in output (default: 80).
- **Returns** — Invisibly returns `output_path`.

```{r export-chroms, eval=FALSE}
# Specific chromosomes
exportChromosomes(genome, c("chr1", "chr22"), "subset.fa")

# All chromosomes
exportChromosomes(genome, output_path = "full.fa")
```

---

## Conversion Utilities

### as_DNAString

```r
as_DNAString(seq_string)
```

Convert a character string to a Biostrings `DNAString` object.

- **seq_string** — Character string containing a DNA sequence.
- **Returns** — A `DNAString` object.
- **Errors** — If Biostrings is not installed.

```{r as-dnastring, eval=FALSE}
dna <- as_DNAString("ACGTACGT")
```

### as_DNAStringSet

```r
as_DNAStringSet(seq_strings, names = NULL)
```

Convert a character vector to a Biostrings `DNAStringSet` object.

- **seq_strings** — Character vector of DNA sequences.
- **names** — Optional names for the sequences.
- **Returns** — A `DNAStringSet` object.
- **Errors** — If Biostrings is not installed.

```{r as-dnastringset, eval=FALSE}
seqs <- as_DNAStringSet(c("ACGT", "GGCC"), names = c("seq1", "seq2"))
```

---

## Working with the Underlying Store

The `store()` accessor gives you access to the full `gtars::RefgetStore` API
for operations not directly exposed by BiocRefgetStore.

```{r store-advanced, eval=FALSE}
s <- store(genome)

# List all aliases in the store
gtars::get_aliases(s)

# Compare two sequence collections
gtars::compare_seqcols(s, digest_a, digest_b)

# Get FHR (FASTA Header Record) metadata
gtars::get_fhr(s, collection_digest(genome))

# Access level 2 data (raw attribute arrays)
level2 <- gtars::get_level2(s, collection_digest(genome))
level2$names      # sequence names
level2$lengths    # sequence lengths
level2$sequences  # sequence digests
```

### show

```r
show(object)
```

Display method for `RefgetGenome`. Prints the number of sequences, collection
digest, and first few sequence names.

```{r show, eval=FALSE}
genome
#> RefgetGenome with 24 sequences
#>   collection_digest: abc123...
#>   seqnames: chr1, chr2, chr3, chr4, chr5 ... (19 more)
```
